---
title: "Exploring Covid-19 Vaccination Progress"
author: "Jane Pratt"
date: "01/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

For this project, we'll be exploring the data about Covid-19 Vaccinations gathered by "Our World In Data". 

## Goals

- To show the total number of people vaccinated worldwide:
  * At the current time
  * Visualizing the total over time
- To show the number of countries reporting milestones over time:
- To show the percentage of the population vaccinated by each country:
  * Listing the current results in a sortable and searchable table
  * Visualizing the current results in a choropleth map
  * Visualizing the current results in a dot density map
  * Visualizing the results for the top n countries over time in an animated bar chart
- To visualize the vaccination progress on a choropleth map
- To explore the potential relationship between  various economic measurements and vaccination rates, for example:
  * GDP and / or GDP per capita
  * World Bank Income Income Groups

## Libraries

```{r, include= FALSE}
library(tidyverse)
```

# Loading the Data

## Downloading the data
We'll be using the Data on COVID-19 (coronavirus) vaccinations from Our World in Data. For details of the data set, see the Further Information section at the end of this document. 

The data is updated daily, so if the file should be downloaded if it either does not exist locally, or has not been modified for more than 1 day.

```{r get_data}
source_location <- 'https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv'
local_file_location <- 'Data/covid_Vaccination_data.csv'
result <- download.file(source_location, destfile = local_file_location, 
                          method="wininet", quiet = TRUE)  
```

## Reading the Local Data

Let's start by reading in the data from the local CSV file, and having a look at the structure of the data:

```{r}
vaccination_data <- read.csv(save_file)
str(vaccination_data)
```



# Further Information

Description of the Our World in Data COVID-19 Vaccination data: 
https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations



